An active sampling model of mental rotation

### 2014-01-27 man/cogsci-2014/cogsci-2014.tex
Work on results section

@@

\input{../../results/D/response_time.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy.tex}
\input{../../results/D/accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/trial_accuracy_corrs.tex}

@@

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/model-traces.pdf}
    \caption{\textbf{Model traces.}}
    \label{fig:model-traces}
  \end{center}
\end{figure*}

@@

\subsubsection{Threshold-based model}

An intuitive and simple model is to first rotate in the direction that
increases similarity, and then stop rotating once a ``match'' is
found, as determined by a threshold on the value of $S$. If all
rotations have been exhausted, then flip, and try rotating again.

@@

\subsubsection{Hill-climbing model}

Unfortunately, it is not always clear what constitutes a good enough
``match''. Many shapes may have multimodal similarity functions
\TODO{refer to figure}, and the maximum possible value of $S$ is
unlikely to be known \textit{a priori} \TODO{cite something about how
  shape matching is hard}. In this case, we must weaken our stopping
criteria: rather than searching for a value above a threshold, we
search for \textit{any} local maximum.  Similar to the threshold-based
model, we use the (local) MLE estimate of $\theta$, as in
Eq.~\ref{eq:mle-lh-ratio}.

\subsubsection{Confidence-based model}

An alternative to the previous two models is to approximate the
integrals in Eq.~\ref{eq:lh-ratio} by picking locations to sample
which will most improve accuracy of the estimate. Rather than a point
estimate of the ratio, or even a simple Monte Carlo estimate, we can
infer a distribution by placing priors over the similarity functions
themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors \cite{Osborne:2012tm} \TODO{probably
  need to describe better what a GP is}. Because our domain is on a
circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
$\log(S_h)$. Because GPs have no guarantee of non-negativity, placing
the prior over the log enforces positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP.\footnote{Note that the approach
  presented in \cite{Osborne:2012tm} differs slightly from the one
  used here. We chose this newer version after a discussion with one
  of the authors on the original BQ paper \TODO{cite personal
    communication}.}We will denote these points as
$\bar{S}_h:=\exp(\mu_h)$. Then, from \TODO{cite personal
  communication}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time.pdf}
    \caption{\textbf{Response times.}}
    \label{fig:response-times}
  \end{center}
\end{figure*}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=0.85\textwidth]{../../figures/D/accuracy.pdf}
    \caption{\textbf{Accuracy.}}
    \label{fig:response-times}
  \end{center}
\end{figure*}

@@

We ran model simulations for the same stimuli that participants
viewed. We ran 10 samples of each stimulus under the Oracle model, and
100 samples of each stimulus for the Threshold, Hill Climbing, and
Bayesian Quadrature models, with the exception of stimuli where
$\theta=0$ or $\theta=180$, in which case we ran double the number of
samples.

For all analyses (except those of accuracy), confidence intervals were
computed using a bootstrap analysis with 10000 bootstrap samples. For
analyses of accuracy, confidence intervals were computed from a
binomial proportion with a Jeffrey's beta prior.

@@

\begin{figure}[t]
  \centering
  \subfloat{\includegraphics[width=0.25\textwidth]{../../figures/D/trial_accuracy.pdf}}%
  \subfloat{\includegraphics[width=0.25\textwidth]{../../figures/D/trial_time.pdf}}
  \caption{\textbf{Effect of learning.}}
  \label{fig:learning}
\end{figure}

response time as a function of rotation

accuracy as a function of rotation

practice effects

@@

>>>>

\newcommand{\Oc}[0]{Oracle}
\newcommand{\Th}[0]{Threshold}
\newcommand{\Hc}[0]{HC}
\newcommand{\Bq}[0]{BQ}

@@

\input{../../results/D/response_time.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy.tex}
\input{../../results/D/accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/theta_time_corrs.tex}
\input{../../results/D/theta_accuracy_corrs.tex}

@@

\TODO{put correlations in figures}

\TODO{need a figure with an example similarity curve, and example
  stimuli}

@@

@@

\subsubsection{\Oc{} model}

The previous hypothesis in the literature is that people perform a
mental rotation in the shortest direction until a match is found, and
that determining this direction is an implementation detail \TODO{cite
  anderson?}. To reflect this hypothesis, we created an ``oracle''
model which knows the correct rotation \textit{a priori}, and simply
performs rotations in the shortest direction until it reaches the
correct rotation.

\subsubsection{\Th{} model}

An intuitive and simple model which does not assume the direction is
known might first rotate in the direction that increases similarity,
and then stop rotating once a ``match'' is found, as determined by a
threshold on the value of $S$. If all rotations have been exhausted,
then flip, and try rotating again.

@@

\subsubsection{Hill climbing model}

Unfortunately, it is not always clear what constitutes a good enough
``match''. Many shapes may have multimodal similarity functions
\TODO{refer to figure}, and the maximum possible value of $S$ is
unlikely to be known \textit{a priori} \TODO{cite something about how
  shape matching is hard}. In this case, we must weaken our stopping
criteria: rather than searching for a value above a threshold, we
search for \textit{any} local maximum, as in the Hill Climbing search
algorithm (\Hc{}). Similar to the \Th{} model, we use the (local) MLE
estimate of $\theta$, as in Eq.~\ref{eq:mle-lh-ratio}.

\subsubsection{Bayesian quadrature model}

An alternative to the previous models is not to perform a search for
the MLE, but to approximate the integrals in Eq.~\ref{eq:lh-ratio} by
picking locations to sample which will most improve accuracy of the
estimate. Rather than a point estimate of the ratio, or even a simple
Monte Carlo estimate, we can infer a distribution by placing priors
over the similarity functions themselves:
\begin{equation}
\ell = \frac{\int \left[\int S(\Xb, f(\Xa, \theta, \hi))p(\theta)\dif\theta\right] p(S)\dif S}{\int \left[\int S(\Xb, f(\Xa, \theta, \hf))p(\theta)\dif\theta\right] p(S)\dif S}
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use Gaussian Process (GP) priors \cite{Osborne:2012tm} \TODO{probably
  need to describe better what a GP is}. Because our domain is on a
circle, we utilize a periodic kernel for the covariance
\cite{Rasmussen:2006vz}, i.e. $k(\theta,
\theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
$\log(S_h)$. Because GPs have no guarantee of non-negativity, placing
the prior over the log enforces positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP.\footnote{Note that the approach
  used here differs slightly from the one presented in
  \citeA{Osborne:2012tm} and is based on personal communication with
  one of the authors \cite{Duvenaud:2013td}.}We will denote these
points as $\bar{S}_h:=\exp(\mu_h)$. Then, from
\citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time.pdf}
    \caption{\textbf{Response times.} \TODO{}}
    \label{fig:response-times}
  \end{center}
\end{figure*}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=0.85\textwidth]{../../figures/D/accuracy.pdf}
    \caption{\textbf{Accuracy.} \TODO{}}
    \label{fig:accuracy}
  \end{center}
\end{figure*}

@@

We ran model simulations for the same stimuli that participants
viewed. We ran 10 samples of each stimulus under the \Oc{} model, and
100 samples of each stimulus for the \Th{}, \Hc{}, and
\Bq{} models, with the exception of stimuli where
$\theta=0$ or $\theta=180$, in which case we ran double the number of
samples.

For analyses of response time, confidence intervals were computed
using a bootstrap analysis (sampling with replacement) with 1000
bootstrap samples. Response times were only only considered for
correct responses.

For analyses of accuracy, confidence intervals were
computed from a binomial proportion with a Jeffrey's beta prior.  For
correlations, we performed a bootstrap analysis (sampling with
replacement) over Spearman correlations using 10000 bootstrap samples.

To test if participants or a model was above chance on a particular
stimulus, we used the same posterior and checked whether $p>0.05$ for
the 50\% mark.

@@

\begin{figure}[t]
  \centering
  \subfloat{\includegraphics[width=0.25\textwidth]{../../figures/D/trial_accuracy.pdf}}%
  \subfloat{\includegraphics[width=0.25\textwidth]{../../figures/D/trial_time.pdf}}
  \caption{\textbf{Effect of learning.} \TODO{}}
  \label{fig:learning}
\end{figure}

\subsubsection{Response times}

The average response time across all stimuli was \ExpTime{} seconds,
and the average response time for an individual stimulus ranged from
\ExpTimeMin{} to \ExpTimeMax{} seconds. The average accuracy across
all stimuli was \ExpAccuracy{}, though there were \ExpNumChance{}
individual stimuli (out of 720) for which people were not above
chance.

As expected, the minimum angle of rotation was significanty
monotonically correlated with average per-stimulus response times,
both for flipped pairs (\ExpThetaTimeCorrFlipped{}) and same pairs
(\ExpThetaTimeCorrSame{}). This was also the case for the number of
actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{}
model (\ThThetaTimeCorr{}), and the \Bq{} model
(\BqThetaTimeCorr{}), but not the \Hc{} model
(\HcThetaTimeCorr{}). Fig.~\ref{fig:response-times}a

The number of actions taken by the \Oc{} model was the best predictor
of human response times, with a correlation of \ExpOcTimeCorr{}. The
\Th{} model was the next best (\ExpThTimeCorr{}), with the \Bq{} model
not far behind (\ExpBqTimeCorr{}). The \Hc{} model was not
significantly correlated with people (\ExpHcTimeCorr{}).

\TODO{flipped response times higher than same response times for
  smaller angles?}

\subsubsection{Accuracy}

Of the models, the \Oc{} and \Th{} models peformed perfectly,
with 100\% accuracy. The \Bq{} model was very accurate
overall (\BqAccuracy{}), though there were \BqNumChance{} stimuli for
which it was not above chance. The \Hc{} model was not
particularly accurate, and although it did perform above chance
overall (\HcAccuracy{}), there were \HcNumChance{} individual stimuli
for which it was not above chance.

The minimum angle was also correlated with the average per-stimulus
accuracy: for flipped pairs, \ExpThetaAccuracyCorrFlipped{}, and for
same pairs, \ExpThetaAccuracyCorrSame{}. There was also a significant
effect for the \Hc{} model (\HcThetaAccuracyCorr{}) and the
\Bq{} model
(\BqThetaAccuracyCorr{}). Fig.~\ref{fig:accuracy}

There was a moderate correlation between the accuracy of the \Hc{}
model and human accuracy (\ExpHcAccuracyCorr{}). The \Bq{} model had a
similar correlation, with \ExpBqAccuracyCorr{}.

\subsubsection{Practice effects}

There was a significant effect of learning both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second
half). Fig.~\ref{fig:learning}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/model-traces.pdf}
    \caption{\textbf{Model traces.} \TODO{} \TODO{put pictures of the
        stimuli here}}
    \label{fig:model-traces}
  \end{center}
\end{figure*}

We replicated the previous monotonic response time result

Also replicated the constant accuracy for flipped stimuli, but not for
same stimuli, from \TODO{cite cooper}. Why is this? Maybe the strategy
is ``they are different until proven the same''.

None of the models were particularly good fits to human data. The best
fit was the \Oc{} model, though this is not satisfying because the
\Oc{} model knows the answer beforehand, which people obviously do
not. The next best model, the \Th{} model, is also not a
particularly satisfying account, because it knows the global maximum
\textit{a priori}, which is not a reasonable assumption. \TODO{explain
  why it is faster for 180 than mid angles -- it is because for a lot
  of stimuli, there is rotational symmetry at 180, so it almost always
  starts checking the correct version first. For mid angles, it might
  check the wrong version first, meaning it takes much longer.}

The \Bq{} model makes more reasonable assumptions about the observer's
knowledge, but has its own problems. Because it is forced to take
small steps, it often gets stuck sampling the same region over and
over again, particularly for stimuli with larger rotations
(Fig.~\ref{fig:model-traces}). As a result, its response time curve is
convex rather than concave.

\TODO{talk about how these results suggest a model of mental rotation
  is harder than was previous thought, and that this is a first step
  towards detangling these issues.}
